Modularly Composing Typed Language Fragments

### 2014-11-04 att-pldi15/att-pldi15.tex
Updated PLDI15

In this paper, we begin from first principles by specifying a language, @$\lambda$, in a style that follows many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show constrained strings and a variation on record types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of its associated term-level \emph{opcons} (e.g. row projection) using a static language where types and translations are values. The semantics guarantees \emph{type safety} and, unlike other approaches, \emph{conservativity}: that all \emph{tycon-specific invariants} that can be established in a ``closed world'' will be conserved in the ``open world''. Type system providers need not provide mechanized proofs. Instead, these follow from a form of translation validation that uses type abstraction to check that tycons maintain \emph{translation independence}, analagous to the representation independence maintained by ML-style module systems.
\end{abstract}

@@

Universality properties (e.g. ``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. For example, in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because you need the row projection operators, in ML written $\opname{\#}\conclbl{lbl}$, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$ itself is an \emph{operator constructor}). %You can only use $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ as a \emph{target of compilation} for a language with record types (they can be implemented using nested binary products,  lists, or other types). %(and attempting to simulate records via a hash table over strings plus a refinement type system to check that the lookup operation will not fail will not satisfy a cost semantics prescribing $\mathcal{O}(1)$ lookup; see Sec. \ref{prior-work}).

%Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a fragment must be introduced directly, rather than as a library, a new \emph{dialect} of the language arises. Within the ML lineage, dialects that go beyond ``core ML'' abound:%, showing that this is not an isolated example:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

@@

Mechanisms are needed that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language. For example, if a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ should continue to behave as strings in the regular language indicated no matter which other extensions are in use by a client. %Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today. %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.

>>>>

In this paper, we begin from first principles by specifying a language, @$\lambda$, in a style that follows many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show constrained strings and a variation on record types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of its associated term-level \emph{opcons} (e.g. row projection) using a static language where types and translations are values. The semantics guarantees \emph{type safety} and, critically, \emph{conservativity}: that all \emph{tycon-specific invariants} that can be established in a ``closed world'' will necessarily be conserved in the ``open world''. Type system providers need not provide mechanized proofs. Instead, these guarantees require only a form of translation validation that uses type abstraction to check that tycons maintain \emph{translation independence}, analagous to representation independence in ML-style module systems.
\end{abstract}

@@

Universality properties (e.g. ``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified by a cost semantics. This is not always possible. For example, in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because you need row projection operators, in ML written $\opname{\#}\conclbl{lbl}$, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$ itself is thus an \emph{operator constructor}). Each time a fragment like must be introduced directly, rather than as a library, a new \emph{dialect} of the language arises. Within the ML lineage, dialects that go beyond ``core ML'' abound:%, showing that this is not an isolated example:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%%You can only use $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ as a \emph{target of compilation} for a language with record types (they can be implemented using nested binary products,  lists, or other types). %(and attempting to simulate records via a hash table over strings plus a refinement type system to check that the lookup operation will not fail will not satisfy a cost semantics prescribing $\mathcal{O}(1)$ lookup; see Sec. \ref{prior-work}).

%Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped.

@@

Mechanisms are needed that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language. For example, if a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{.+}}$ should continue to behave as non-empty strings no matter which other extensions are in use. %Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today. %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.
